---
title: "Finnish personal ID number data toolkit for R (hetu)"
author: "Pyry Kantanen, Jussi Paananen, Mans Magnusson, Leo Lahti"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
  rmarkdown::md_vignette:
    toc: true   
vignette: >
  %\VignetteIndexEntry{Finnish personal ID number data toolkit for R (hetu)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The **hetu R package** provides tools to work with Finnish personal identity numbers.

Where possible, we have unified the syntax with [sweidnumbr](https://github.com/rOpenGov/sweidnumbr).



## Installation

Install the current devel version in R:

```{r install, eval=FALSE}
devtools::install_github("ropengov/hetu")
```

Test the installation by loading the library:

```{r test, message=FALSE, warning=FALSE, eval=TRUE}
library(hetu)
```

We also recommend setting the UTF-8 encoding:

```{r locale, eval=FALSE}
Sys.setlocale(locale="UTF-8") 
```

## Personal identification number (HETU)

The basic hetu function can be used to view information from a Finnish personal identification number. The data is outputted as a data frame.

```{r hetu_example1, message=FALSE}
example_pin <- "111111-111C"
hetu(example_pin)
```

The output can be made prettier, for example by using knitr:

```{r hetu_example2, message=FALSE}
knitr::kable(hetu(example_pin))
```

The hetu function also accepts vectors with several identification numbers as input:

```{r hetu_example3, message=FALSE}
example_pins <- c("010101-0101", "111111-111C")
knitr::kable(hetu(example_pins))
```

The hetu function prints warning messages to the user if input vector contains invalid PINs.

```{r hetu_example4, error = TRUE, purl = FALSE}
hetu(c("010101-0102", "111311-111C", "010101-0101"))
```

The validity of the PINs can also be determined by using the pin_ctrl function:

```{r pin_ctrl_example1, fig.message=FALSE}
pin_ctrl("010101-0101") # TRUE/FALSE 
pin_ctrl("010101-1010") # FALSE
```

### Extracting specific information

Information contained in the PIN can be extracted with a generic extract parameter. Valid values for extraction are *hetu*, *sex*, *personal.number*, *checksum*, *date*, *day*, *month*, *year*, *century* and *is.temp*.

```{r hetuextract1, message=FALSE}
hetu(example_pins, extract = "sex")
hetu(example_pins, extract = "checksum")
```

Some fields can be extracted with specialized functions. Extracting sex with pin_sex function:

```{r hetuextract2, message=FALSE, eval=TRUE}
pin_sex(example_pins)
```

Extracting age at current date and at a given date with pin_age function:

```{r hetuextract3, message=FALSE, eval=TRUE}
pin_age(example_pins)
pin_age(example_pins, date = "2012-01-01")
```

Dates (birth dates) also have their own function, pin_date.

```{r hetuextract4, message=FALSE, eval=TRUE}
pin_date(example_pins)
```

### Artificial and temporary personal identification numbers

The package functions can be made to accept artificial or temporary personal identification numbers. These temporary PINs are sometimes used in different organizations, such as insurance companies or hospitals, if the exact identity of the individual is not known. Artificial or temporary PINs are not intended for continuous, long term use and they are not usually accepted by PIN validity checking algorithms.

Temporary PINs provide similar information about individual's birthdate or sex as regular PINs. Temporary PINs can also be safely used for testing purposes, as such a number cannot be any person's PIN.

Artificial and temporary PINs can be used normally by allowing them through allow.temp parameter.

```{r hetu_temp1, message=FALSE}
example_temp_pin <- "010101A900R"
knitr::kable(hetu(example_temp_pin, allow.temp = TRUE))
```

A vector with regular and temporary PINs mixed together prints only regular PINs, if allow.temp is not set to TRUE. Automatic omitting of temporary PINs does not produce a visible error message and therefore users need to be cautious if they want to use temporary PINs.

If temporary PINs are not explicitly allowed and the input vector consists of temporary PINs only, the function will return an error.

```{r hetu_temp2, error = TRUE, purl = FALSE}
example_temp_pins <- c("010101A900R", "010101-0101")
knitr::kable(hetu(example_temp_pins))
```

When allow.temp is set to TRUE, all PINs are handled as if they were regular PINs.

```{r hetu_temp3, message=FALSE}
knitr::kable(hetu(example_temp_pins, allow.temp = TRUE))
```

Validation function pin_ctrl produces a FALSE for every artificial / temporary PIN, if they are not explicitly allowed.

```{r hetu_temp4, error = TRUE, purl = FALSE}
knitr::kable(hetu(example_temp_pins)) #FALSE TRUE
knitr::kable(hetu(example_temp_pins, allow.temp = TRUE)) #TRUE TRUE
```

## Generating random PINs

Random PINs can be generated by using the rpin function.

```{r rpin, message=FALSE}
rpin(n = 4)
rpin(n = 4, start_date = "1990-01-01", end_date = "2005-01-01")
```

The number of males in the generated sample can be changed with parameter p.male. Default is 0.4.

```{r rpin2, message=FALSE}
random_sample <- rpin(n = 4, p.male = 0.8)
table(random_sample)
```

The default proportion of artificial / temporary PINs is 0.0, meaning that no artificial / temporary PINs are generated by default.

```{r rpin3, message=FALSE}
temp_sample <- rpin(n = 4, p.temp = 0.5)
table(hetu(temp_sample, allow.temp = TRUE, extract = "is.temp"))
```

## Various examples

A simple example on how to use hetu package in tidyverse/dplyr manner to extract information to tibble format.

```{r hetu_tibbles, eval = FALSE}
library(hetu)
library(tidyverse)
library(dplyr)

# Generate data for this example
hdat<-tibble(pin=rpin(n = 4, start_date = "1990-01-01", end_date = "2005-01-01"))

# Extract all the hetu information to tibble format
hdat<-hdat %>%
  mutate(result=map(.x=pin,.f=hetu::hetu)) %>% unnest(cols=c(result))
hdat
```

```{r example2, message=FALSE, echo=FALSE, eval=FALSE}
# sweidnumbr additionally has class structure for IDs
#example_pin <- as.pin(example_pin)
#example_pin

#The next step is to test if the vector is a ```pin``` object. To do
#this we use the ```is.pin()``` function.
#is.pin(example_pin)

#This only check the format of the pin. To check the pin using the
#control number we use ```pin_ctrl()```.
# pin_ctrl(example_pin)

#It is also possible to format the pin for presentation in different forms. (Note however that #the output of `format_pin` is just a character and no longer a `pin` object):

#format_pin(example_pin, "%Y-%m-%d-%N")
#format_pin(example_pin, "%P")

#Sometimes we want some example `pin`s. We can easily simulate `pin`s using `rpin()`:
#rpin(3)

```



## Licensing and Citations

This work can be freely used, modified and distributed under the open license specified in the [DESCRIPTION file](https://github.com/MansMeg/hetu/blob/master/DESCRIPTION).

Kindly cite the work as follows

```{r citation, message=FALSE, eval=TRUE}
citation("hetu")
```

## References

- [The personal identity code](https://dvv.fi/en/personal-identity-code). Digital and population data services agency.
- [Valtioneuvoston asetus väestötietojärjestelmästä (128/2010)](https://www.finlex.fi/fi/laki/ajantasa/2010/20100128). Valtiovarainministeriö.
- [HETU-uudistuksen loppuraportti](http://urn.fi/URN:ISBN:978-952-367-296-3). Valtiovarainministeriön julkaisuja 2020:20.

## Session info

This vignette was created with

```{r sessioninfo, message=FALSE, warning=FALSE}
sessionInfo()
```

