---
title: "hetu : Structural handling of Finnish identity numbers"
author: "Leo Lahti, Jussi Paananen, Mans Magnusson"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
  rmarkdown::md_vignette:
    toc: true   
vignette: >
  %\VignetteIndexEntry{hetu}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The **hetu R package** provides tools to work with Finnish personal identity numbers.

Where possible, we have unified the syntax with [sweidnumbr](https://github.com/rOpenGov/sweidnumbr).



## Installation

Install the current devel version in R:

```{r install, eval=FALSE}
devtools::install_github("ropengov/hetu", ref = "hetu-devel")
```

Test the installation by loading the library:

```{r test, message=FALSE, warning=FALSE, eval=TRUE}
library(hetu)
```

We also recommend setting the UTF-8 encoding:

```{r locale, eval=FALSE}
Sys.setlocale(locale="UTF-8") 
```


## Personal identification number (HETU)

At simplest, we can use the following to extract information from a Finnish personal identification number.

```{r example, message=FALSE}
hetu("111111-111C")
```

The function accepts also vectors as input, returning a data frame:

```{r hetuvec, message=FALSE}
example_pin <- c("010101-0101", "111111-111C")
knitr::kable(hetu(example_pin))
```

Validating Finnish personal identification number:

```{r hetu2, fig.message=FALSE}
pin_ctrl("010101-0101") # TRUE/FALSE 
```

### Artificial and temporary personal identification numbers

The package functions have the option to accept artificial or temporary personal identification numbers. These numbers are sometimes used internally in different organizations, such as insurance companies or hospitals, if the exact identity of the individual is not known. Artificial or temporary PINs are not intended for continuous or external use and are not usually accepted by validity checking algorithms.

However, even temporary PINs provide similar information about individual's birthdate or sex as regular PINs. Temporary PINs can also be safely used for testing purposes.

```{r hetu_temp1, message=FALSE}
hetu("010101A900R") #NA
hetu("010101A900R", allow.temp = TRUE)
```

Even one temporary pin can cause the functions to produce an error. 

```{r hetu_temp2, error = TRUE, purl = FALSE}
hetu(c("010101A900R", "010101-0101")) #Error
hetu(c("010101A900R", "010101-0101"), allow.temp = TRUE) #TRUE
```

Validation function \code{pin_ctrl} produces a \code{FALSE} for every artificial / temporary pin, if they are not explicitly allowed.

```{r hetu_temp3, message=FALSE}
pin_ctrl(c("010101A900R", "010101-0101")) #FALSE TRUE
pin_ctrl(c("010101A900R", "010101-0101"), allow.temp = TRUE) #TRUE TRUE
```

### Extracting specific fields

General field extraction from personal identification number.

```{r hetuextract, message=FALSE}
hetu(example_pin, extract = "sex")
hetu(example_pin, extract = "checksum")
```

Extracting sex with a specific function:

```{r example4, message=FALSE, eval=TRUE}
pin_sex(example_pin)
```

Extracting age at current date and at a given date:

```{r example5, message=FALSE, eval=TRUE}
pin_age(example_pin)
pin_age(example_pin, date = "2012-01-01")
```

## Generating random PINs

Random PINs can be generated by using the rpin function.

```{r rpin, message=FALSE}
rpin(n = 4)
rpin(n = 4, start_date = "1990-01-01", end_date = "2005-01-01")
```

The number of males in the generated sample can be changed with parameter p.male.

```{r rpin2, message=FALSE}
random_sample <- rpin(n = 4, p.male = 0.8)
table(random_sample)
```

The default proportion of artificial / temporary pins is 0, meaning that no artificial / temporary PINs are generated at default.

```{r rpin3, message=FALSE}
temp_sample <- rpin(n = 4, p.temp = 0.5)
table(hetu(temp_sample, allow.temp = TRUE, extract = "is.temp"))
```

```{r example2, message=FALSE, echo=FALSE, eval=FALSE}
# sweidnumbr additionally has class structure for IDs
#example_pin <- as.pin(example_pin)
#example_pin

#The next step is to test if the vector is a ```pin``` object. To do
#this we use the ```is.pin()``` function.
#is.pin(example_pin)

#This only check the format of the pin. To check the pin using the
#control number we use ```pin_ctrl()```.
# pin_ctrl(example_pin)

#It is also possible to format the pin for presentation in different forms. (Note however that #the output of `format_pin` is just a character and no longer a `pin` object):

#format_pin(example_pin, "%Y-%m-%d-%N")
#format_pin(example_pin, "%P")

#Sometimes we want some example `pin`s. We can easily simulate `pin`s using `rpin()`:
#rpin(3)

```



## Licensing and Citations

This work can be freely used, modified and distributed under the open license specified in the [DESCRIPTION file](https://github.com/MansMeg/hetu/blob/master/DESCRIPTION).

Kindly cite the work as follows

```{r citation, message=FALSE, eval=TRUE}
citation("hetu")
```

## References

- [The personal identity code](https://dvv.fi/en/personal-identity-code). Digital and population data services agency.
- [Valtioneuvoston asetus väestötietojärjestelmästä (128/2010)](https://www.finlex.fi/fi/laki/ajantasa/2010/20100128). Valtiovarainministeriö.
- [HETU-uudistuksen loppuraportti](http://urn.fi/URN:ISBN:978-952-367-296-3). Valtiovarainministeriön julkaisuja 2020:20.

## Session info

This vignette was created with

```{r sessioninfo, message=FALSE, warning=FALSE}
sessionInfo()
```

