---
title: "Structural handling of Finnish identity numbers (hetu R package)"
author: "Leo Lahti, Jussi Paananen, Mans Magnusson"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
  rmarkdown::md_vignette:
    toc: true    
vignette: >
  %\VignetteIndexEntry{hetu markdown vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{Cairo}
  \usepackage[utf8]{inputenc}  
---

The **hetu R package** provides tools to work with Finnish personal identity numbers.

Where possible, we have unified the syntax with [sweidnumbr](https://github.com/rOpenGov/sweidnumbr).



## Installation

Install the current devel version in R:

```{r install, eval=FALSE}
devtools::install_github("ropengov/hetu")
```

Test the installation by loading the library:

```{r test, message=FALSE, warning=FALSE, eval=TRUE}
library(hetu)
```

We also recommend setting the UTF-8 encoding:

```{r locale, eval=FALSE}
Sys.setlocale(locale="UTF-8") 
```


## Personal identification number (HETU)

At simplest, we can use the following to extract information from a Finnish personal identification number.

```{r example, message=FALSE}
hetu("111111-111C")
```

The function accepts also vectors as input, returning a data frame:

```{r hetuvec, message=FALSE}
example_pin <- c("010101-0101", "111111-111C")
knitr::kable(hetu(example_pin))
```

Validating Finnish personal identification number:

```{r hetu2, fig.message=FALSE}
pin_ctrl("010101-0101") # TRUE/FALSE 
```


### Extracting specific fields

General field extraction from personal identification number.

```{r hetuextract, message=FALSE}
hetu(example_pin, extract = "gender")
```

Extracting sex:

```{r example4, message=FALSE, eval=TRUE}
pin_sex(example_pin)
```


Extracting age:

```{r example5, message=FALSE, eval=TRUE}
pin_age(example_pin)
pin_age(example_pin, date = "2020-01-01")
```


```{r example2, message=FALSE, echo=FALSE, eval=FALSE}
# sweidnumbr additionally has class structure for IDs
#example_pin <- as.pin(example_pin)
#example_pin

#The next step is to test if the vector is a ```pin``` object. To do
#this we use the ```is.pin()``` function.
#is.pin(example_pin)

#This only check the format of the pin. To check the pin using the
#control number we use ```pin_ctrl()```.
# pin_ctrl(example_pin)

#It is also possible to format the pin for presentation in different forms. (Note however that #the output of `format_pin` is just a character and no longer a `pin` object):

#format_pin(example_pin, "%Y-%m-%d-%N")
#format_pin(example_pin, "%P")

#Sometimes we want some example `pin`s. We can easily simulate `pin`s using `rpin()`:
#rpin(3)

```



## Licensing and Citations

This work can be freely used, modified and distributed under the open license specified in the [DESCRIPTION file](https://github.com/MansMeg/hetu/blob/master/DESCRIPTION).

Kindly cite the work as follows

```{r citation, message=FALSE, eval=TRUE}
citation("hetu")
```



## Session info

This vignette was created with

```{r sessioninfo, message=FALSE, warning=FALSE}
sessionInfo()
```
